Exploring Relationships between Demographic Characteristics, Health, Finances, Living Arrangements, and Happiness: An Analysis of the General Social Survey (GSS) Data from 1972 to 2012

### 2023-03-29 outputs/paper/paper.qmd
Add files via upload

Used some R language packages, such as tidyverse [@ti], dplyr [@dplyr], janitor [@janitor] , knitr [@knitr], here [@here], haven [@haven], ggplot [@ggplot] , kableExtra [@kableExtra] to assist in data analysis, and the data from General Social Survey [@gss].

@@

```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(dplyr)
library(tidyr)
library(kableExtra)
library(janitor)
library(knitr)
library(haven)
library(here)
library(ggplot2)
library(tinytex)
data <- read_csv(here::here("inputs/data/cleaned_data.csv"))
```

@@

data %>%
  select(-1:-8) %>%
  slice(1:5) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   year: This variable indicates the year in which the data was collected. In this data-set, all the observations are from the year 1972 to 2012.
-   age: This variable indicates the age of the individual at the time the data was collected. It is a continuous variable measured in years.
-   sex: This variable indicates the biological sex of the individual. It is a categorical variable with two levels: male and female.
-   babies: This variable indicates the number of babies the individual had at the time the data was collected. It is a categorical variable with different levels, depending on the number of babies: 0, 1, 2, 3, 4, or 5+.
-   divorce: This variable indicates whether or not the individual had ever been divorced at the time the data was collected. It is a categorical variable with two levels: yes and no.
-   health: This variable indicates the self-reported health status of the individual at the time the data was collected. It is a categorical variable with different levels, depending on the health status: excellent, good, fair, or poor.
-   finance_satisfied: This variable indicates the individual's level of satisfaction with their current financial situation. It is a categorical variable with different levels of satisfaction: very satisfied, more or less satisfied, not very satisfied, and not at all satisfied.
-   living_area: This variable indicates the type of living area in which the individual resided at the time the data was collected. It is a categorical variable with different levels: large city, small city, suburban, and rural.
-   happy: This variable indicates the individual's level of happiness or life satisfaction. It is a categorical variable with different levels of happiness: very happy, pretty happy, not too happy, and not at all happy.

@@

```{r}
#| warning: false
#| echo: false
# create a new variable "divorce_score" based on the "divorce" variable
data$divorce_score <- ifelse(data$divorce == "no", 0, 1)

data$happy_score <- ifelse(data$happy == "not too happy", 1,
                      ifelse(data$happy == "not satisfied at all", 2,
                      ifelse(data$happy == "more or less satisfied", 3,
                      ifelse(data$happy == "pretty well satisfied", 4, 5))))

# Create ordinal variable for degree score
data$degree_score <- ifelse(data$degree == "Less than high school", 1,
                            ifelse(data$degree == "High school graduate", 2,
                                   ifelse(data$degree == "Some college", 3,
                                          ifelse(data$degree == "Bachelor's degree", 4,
                                                 ifelse(data$degree == "Advanced degree", 5, NA)))))

# Create ordinal variable for health score
data$health_score <- ifelse(data$health == "Poor", 1,
                            ifelse(data$health == "Fair", 2,
                                   ifelse(data$health == "Good", 3,
                                          ifelse(data$health == "Very good", 4,
                                                 ifelse(data$health == "Excellent", 5, NA)))))

# Create ordinal variable for finance score
data$finance_score <- ifelse(data$finance_satified == "Not at all satisfied", 1,
                            ifelse(data$finance_satified == "Not too satisfied", 2,
                                   ifelse(data$finance_satified == "More or less satisfied", 3,
                                          ifelse(data$finance_satified == "Pretty well satisfied", 4,
                                                 ifelse(data$finance_satified == "Completely satisfied", 5, NA)))))

# convert age variable from character to numeric
data$age <- as.numeric(data$age)

# perform linear regression of age and divorce_score
model <- lm(happy_score ~ age +  degree + finance_satified + health + divorce, data=data)
model
```

>>>>

Used some R language packages, such as tidyverse [@ti], dplyr [@dplyr], janitor [@janitor] , knitr [@knitr], here [@here], haven [@haven], ggplot [@ggplot] , kableExtra [@kableExtra] , model summary [@modelsummary] to assist in data analysis, and the data from General Social Survey [@gss].

@@

```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(dplyr)
library(tidyr)
library(kableExtra)
library(janitor)
library(knitr)
library(haven)
library(here)
library(ggplot2)
library(tinytex)
library(modelsummary)
data <- read_csv(here::here("inputs/data/cleaned_data.csv"))
```

@@

data %>%
  select(9:13) %>%
  slice(1:5) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
#| eval: true
#| warning: false
#| echo: false
data %>%
  select(14:16) %>%
  slice(1:5) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   year: This variable indicates the year in which the data was collected. In this data-set, all the observations are from the year 1972 to 2012.
-   age: This variable indicates the age of the individual at the time the data was collected. It is a continuous variable measured in years.
-   sex: This variable indicates the biological sex of the individual. It is a categorical variable with two levels: male and female.
-   babies: This variable indicates the number of babies the individual had at the time the data was collected. It is a categorical variable with different levels, depending on the number of babies: 0, 1, 2, 3, 4, or 5+.
-   divorce: This variable indicates whether or not the individual had ever been divorced at the time the data was collected. It is a categorical variable with two levels: yes and no.
-   health: This variable indicates the self-reported health status of the individual at the time the data was collected. It is a categorical variable with different levels, depending on the health status: excellent, good, fair, or poor.
-   finance_satisfied: This variable indicates the individual's level of satisfaction with their current financial situation. It is a categorical variable with different levels of satisfaction: very satisfied, more or less satisfied, not very satisfied, and not at all satisfied.
-   living_area: This variable indicates the type of living area in which the individual resided at the time the data was collected. It is a categorical variable with different levels: large city, small city, suburban, and rural.
-   happy: This variable indicates the individual's level of happiness or life satisfaction. It is a categorical variable with different levels of happiness: very happy, pretty happy, not too happy, and not at all happy.
-   divorce_score: This variable is a binary representation of the divorce status of an individual in the dataset. The value '0' denotes that the individual is not divorced, while the value '1' indicates that they are divorced.
-   happy_score: The happy_score variable is an ordinal variable representing the self-reported happiness levels of the individuals in the dataset. It ranges from 0 to 3, with 0 representing "not too happy," 1 representing "happy," 2 representing "very happy," and 3 reserved for any unclassified or missing responses.
-   degree_score: This ordinal variable represents the educational attainment of the individuals in the dataset. It ranges from 1 to 5, where 1 signifies "less than high school," 2 denotes "high school," 3 represents "associate/junior college," 4 refers to "bachelor's," and 5 indicates "graduate" level education. The value 6 is reserved for any unclassified or missing responses.
-   health_score: The health_score variable is an ordinal variable indicating the self-reported health status of the individuals in the dataset. It ranges from 1 to 4, with 1 denoting "poor" health, 2 signifying "fair" health, 3 representing "good" health, and 4 referring to "excellent" health. The value 5 is reserved for any unclassified or missing responses.
-   finance_score: This ordinal variable represents the financial satisfaction level of the individuals in the dataset. It ranges from 1 to 3, with 1 indicating "Not at all satisfied," 2 denoting "more or less satisfied," and 3 signifying "pretty well satisfied." The value 4 is reserved for any unclassified or missing responses.
-   babies_number: The babies_number variable is an ordinal variable denoting the number of babies an individual in the dataset has. It ranges from 0 to 4, where 0 represents "0" babies, 1 denotes "1" baby, 2 signifies "2" babies, 3 indicates "3" babies, and 4 refers to "4" babies. The value 5 is reserved for any unclassified or missing responses.

@@

$$
\hat{Y}=\hat{\beta_0}+\hat{\beta_1}X_{age}+\hat{\beta_2}X_{degree-score}+\hat{\beta_3}X_{finance-score}+\hat{\beta_4}X_{health-score} + \hat{\beta_5}X_{divorce-score} + 
\hat{\beta_6}X_{babies-number}
$$ {#eq-model}

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-lm
#| tbl-cap: Linear Model and its Summary Statistics
model <- lm(data$happy_score ~ age +  degree_score + finance_score + health_score + divorce_score + babies_number, data=data)

modelsummary :: modelsummary(
  list(
    "Well-Being" = model
  ),
  fmt = 2,
  statistic = "conf.int"
)
```

The [@tbl-lm], which is based the equation of @eq-model :

-   Overall model significance: The F-statistic (314.461) indicates that the overall model is statistically significant, suggesting that the predictors included in the model collectively explain a meaningful portion of the variance in well-being.
-   Model fit: The R-squared (0.070) and adjusted R-squared (0.070) values are relatively low, indicating that only 7% of the variation in well-being is explained by the predictors in the model. This suggests that there may be other factors influencing well-being that have not been included in the model.
-   Age as a predictor: The coefficient for age (0.00) and its confidence interval (\[0.00, 0.00\]) suggest that age does not have a statistically significant effect on well-being in this model.
-   Degree score as a predictor: The positive coefficient for degree_score (0.02) and its confidence interval (\[0.01, 0.03\]) indicate that higher degree scores are associated with higher well-being. This suggests a positive relationship between educational achievement and well-being.
-   Finance score as a predictor: The negative coefficient for finance_score (−0.09) and its confidence interval (\[−0.11, −0.08\]) demonstrate that higher finance scores are associated with lower well-being. This indicates a negative relationship between financial stress and well-being.
-   Health score as a predictor: The positive coefficient for health_score (0.32) and its confidence interval (\[0.30, 0.34\]) show that higher health scores are associated with higher well-being. This suggests a strong positive relationship between health and well-being.
-   Divorce score as a predictor: The coefficient for divorce_score (−0.02) and its confidence interval (\[−0.05, 0.01\]) indicate that the effect of divorce_score on well-being is not statistically significant. This suggests that the relationship between divorce and well-being is inconclusive in this model.
-   Number of babies as a predictor: The negative coefficient for babies_number (−0.03) and its confidence interval (\[−0.05, −0.01\]) demonstrate that having more babies is associated with lower well-being. This indicates a negative relationship between the number of children and well-being.

In conclusion of this result, the model shows that degree_score, finance_score, health_score, and babies_number are statistically significant predictors of well-being, while age and divorce_score are not. However, the low R-squared value suggests that there are likely other factors affecting well-being that have not been included in the model.
